Synthesizing Quotient Lenses

### 2018-02-23 papers/QuotientLenses/main.tex
section 2 editing

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:

@@

(* synthesize a lens that maps the comma representation to the space representation *)
let space_to_comma = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}

(* squash QRE maps comma_name to space_name *)
let name = squash comma_name $\to$ space_name using (get space_to_comma)

@@

\subsection{QRE Lenses}
While QREs make it easier to write equivalence relations on regular expressions
by specifying the data and the equivalence relation on the data in one
expression, in order to define a quotient lens, the programmer still has to
manually derive the underlying lens that maps the equivalence classes of the
source data to the equivalence classes of the target data. This is itself a
tricky task as Figure~\ref{fig:example-lens} illustrates.

@@

The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
EndNote data  (and vice versa) by matching the delimiters and separators in a
bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
programmer is writing the lens, she has to make sure that all of these small
deletions and insertions are correct, which can be a challenging task if the
lens is large enough, as in the definition of the lens \verb!author_lens! in
Figure~\ref{fig:example-lens}. This makes lens programming tedious and
error-prone.

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and show how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, which are bijections created
from the Boomerang lens combinators.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs---every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

One potential pitfall with QRE lenses is that the combinators that we define on
them may not preserve this normal form. For example, the composition $q'; q$ of
QRE lenses $q$ and $q'$ might not be equivalent to any quotient lens that has
the canonical normal form of a bijective lens with canonizers at the edges.
Our main technical contribution in this paper is a proof that each QRE lenses
$q$ defines the same transformation as a QRE lens $q'$ that is in normal form.

For example, suppose \cd{q_1 : bibtex} $\Leftrightarrow$ \cd{endnote} is a QRE
lens that maps between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote
<=> xml} is a QRE lens that maps between EndNote and XML formats. The
composition \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} of \cd{q_1} and
\cd{q_2} is not necessarily in normal form due to the presence of canonizers
on each side of the intermediary EndNote format:

\sam{TODO: We need to put a diagram of composition here.}

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml}
of the same type as \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} such
that \cd{q_1 ; q_2} and \cd{q} define the same transformation. This proof
will allow justify our approach for synthesizing quotient lenses so that the
programmer can replace all of the code in Figure~\ref{fig:example-lens} with a
call to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote)
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the resulting  quotient lens maps between the \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_example} to
the equivalence class of \cd{end_example}, which are the two concrete example
strings given at the beginning of this Section.

As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
create lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

>>>>

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?}

@@

(* synthesize a lens that maps the comma representation to the space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l

(* squash QRE maps comma_name to space_name *)
let name = squash comma_name $\to$ space_name using s_to_c

@@

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
quotient lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
\item \kw{del} $s$: delete the constant string $s$ in the left-to-right direction; produce the constant string $s$ in the right-to-left direction.
\item \kw{ins} $s$: produce the string $s$ in the left-to-right direction;
delete the string $s$ in the right-to-left direction.
\item \kw{id} $r$: copy the text matching $r$ from left-to-right (also from
right-to-left) \dpw{this was not the exact syntax.  fix this.}
\end{itemize}
%
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

%% While QREs make it easier to write equivalence relations on regular expressions
%% by specifying the data and the equivalence relation on the data in one
%% expression, in order to define a quotient lens, the programmer still has to
%% manually derive the underlying lens that maps the equivalence classes of the
%% source data to the equivalence classes of the target data. This is itself a
%% tricky task as Figure~\ref{fig:example-lens} illustrates.

@@

%% The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
%% EndNote data  (and vice versa) by matching the delimiters and separators in a
%% bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
%% preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
%% programmer is writing the lens, she has to make sure that all of these small
%% deletions and insertions are correct, which can be a challenging task if the
%% lens is large enough, as in the definition of the lens \verb!author_lens! in
%% Figure~\ref{fig:example-lens}. This makes lens programming tedious and
%% error-prone.

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and by showing how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

Intuitively, our QRE lenses are bijective lenses with ``canonizers at the
edges''---Figure~\ref{fig:attheedges} presents the architecture of each lens.
Every QRE lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

Because the QREs $c$ and $c'$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c$ and $c'$ are all that is required to synthesize
QRE lenses end-to-end.  Indeed, we structured QRE lenses as bijective lenses with
canonizers at the edges, precisely to facilitate the use of this synthesis algorithm.

@@

%% One potential pitfall with QRE lenses is that the combinators that we define on
%% them may not preserve this normal form. For example, the composition $q'; q$ of
%% QRE lenses $q$ and $q'$ might not be equivalent to any quotient lens that has
%% the canonical normal form of a bijective lens with canonizers at the edges.

On the other hand, our restriction that canonizers appear only at the edges
raises a key technical question:  Are we limiting the expressiveness of our transformations
by demanding all programs fit into this normal form?  It turns out that we are not:
any lens that uses canonizers internally can be transformed into a lens that uses canonizers
at the edges.  The main technical contribution of this paper is a proof of this fact.  The
crucial step in the proof is a demonstration that functional composition of two lenses
with ``canonizers at the edge'' can be represented as some single lens with ``canonizers at the
edge.''

%% For example, suppose \cd{q_1 : bibtex} $\Leftrightarrow$ \cd{endnote} is a QRE
%% lens that maps between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote
%% <=> xml} is a QRE lens that maps between EndNote and XML formats. The
%% composition \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} of \cd{q_1} and
%% \cd{q_2} is not necessarily in normal form due to the presence of canonizers
%% on each side of the intermediary EndNote format:

%% \sam{TODO: We need to put a diagram of composition here.}

%% \noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
%% shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml}
%% of the same type as \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} such
%% that \cd{q_1 ; q_2} and \cd{q} define the same transformation.

This technical investigation justifies replacing manually constructed lenses that may use
functional composition and nested canonizers with synthesized lenses in normal form.
The end result is that users may synthesize complex transformations with a single line
of code.  For instance, all the code in Figure~\ref{fig:example-lens} may be replaced by
a single call to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote)
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats.  It uses \cd{bib_example} and \cd{end_example} (the two concrete example
strings given at the beginning of this section) to disambiguate.
In additiona, and as we saw earlier with the definition of \cd{s_to_c}, the synthesis procedure itself
can be used to create lenses that are in turn used to define other QREs.  The ability to intermix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.
